---
title: "Data sets for management"
format: html
editor: visual
---

## Data sets

All data sets will come from the `FSAdata` package. You can consult the package to get more information on the datasets.

\`\`\`{}

\`\`\`{}

You can use the `?` to obtain more info on a particular dataset.

```{r eval=FALSE}
library(FSAdata)
?BGHRfish
```

## Data set 1

#### Big Hill Reservoir

Fish information from electrofishing samples collected from Big Hill Reservoir, KS, 2014.

A dataframe with 266 observations from Smallmouth bass, Largemouth bass and Bluegill.

There are two datasets for this dataset `BGHRfish` and `BGHRsample`.

You are tasked with managing this fishery, and some potential issues are condition, community composition, balance of the fishery and whether this fishery will attract anglers.

## Data set 2

#### InchLake, WI

Make a management plan for Inch Lake, WI. The appropriate datasets are: `BluegillIL`, `InchLake1`, and `InchLake2`. Assume all data is current (that it was collected in the last 2-3 years.

## Data set 3

#### Trout Lake, WI

You have two datasets, each spanning 25 years:

1.  Lengths and weights for Yellow Perch from Trout Lake, WI. `YPerchTL`
2.  Lengths, weights, and sex of Cisco from Trout Lake, WI. `CiscoTL`

Check whether the condition, or size distribution has changed during these 25 years.

## Dataset 4

This is a bit different. You won't be asked to manage a specific body of water, but to manage for Muskellunge abundance in Wisconsin Lakes.

The dataset is `MuskieWI06MR`

```{r eval=F}

```
